---
title: "README"
format: gfm
---
<!-- README.md is generated from README.qmd. Please edit that file -->

# azmet-qaqc

<!-- badges: start -->
[![targets-workflow](https://github.com/cct-datascience/azmet-qaqc/actions/workflows/targets.yaml/badge.svg)](https://github.com/cct-datascience/azmet-qaqc/actions/workflows/targets.yaml)
<!-- badges: end -->

This repository contains code to generate a dataset for doing timeseries forecast-based quality assurance checks for  [AZMET](https://ag.arizona.edu/azmet/) meterological data.  Daily resolution data is accessed via the [`azmetr`](https://github.com/cct-datascience/azmetr) R package. A timeseries model is fit to all but the last day of data and a forecast is made for the last day. This is then uploaded to a Posit Connect server and used downstream by a [validation report](https://github.com/cct-datascience/azmet-qa-dashboard).

## Reproducibility

This project uses [`renv`](https://rstudio.github.io/renv/articles/renv.html) for package management. When opening this repo as an RStudio Project for the first time, `renv` should automatically install itself and prompt you to run `renv::restore()` to install all package dependencies.

This project uses the [`targets` package](https://docs.ropensci.org/targets/) for workflow management. Run `targets::tar_make()` from the console to run the workflow and reproduce all results.  To reproduce the last step that updates the forecast dataset "pin" on Connect, you'll need to create a .Renviron file with 

```
CONNECT_SERVER="https://viz.datascience.arizona.edu/"
CONNECT_API_KEY=<your API key>
```

And you must have permission to publish to that pin.  Without this setup, other targets should still run and you can inspect results of each step with `tar_read()`.

**Targets workflow:**

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
cat(
  "```mermaid",
  targets::tar_mermaid(targets_only = FALSE),
  "```",
  sep = "\n"
)
#this should display correctly on GitHub, or code can be pasted into https://mermaid.live
```


## Repo Structure

```{r}
fs::dir_tree(recurse = 1)
```
- `R/` contains functions used in the `targets` pipeline.
- `_targets` is generated by `targets::tar_make()` and only the metadata of the targets pipeline is on GitHub.
- `_targets.R` defines a `targets` workflow
- `_targets_packages.R` is generated by `targets::tar_renv()`
- `data/` contains the .csv file of the historic data (since 2003)
- `notes/` contains quarto documents of notes / analysis.  `sliding-windows.qmd` is an exploration of a sliding-window quantile approach that is probably not going to work for us.  `azmet-qaqc.qmd` is an exploration of QA by forecasting.
- `renv/` and `renv.lock` are necessary for the `renv` package to work (see above)
- `run.R` is for conveniently running `tar_make()` as a background job.  Created by `targets::use_targets()`
- `wrangling/` contains a script for scraping weather data from 2003 onward that isn't available through the API / the `azmetr` package. This should only need to be run once ever.


## Collaboration guidelines

To contribute to this project, please create a new branch for your changes and make a pull request. One easy way to do this from within R is with the `usethis` package and the `pr_*` functions. `pr_init("branch-name")` begins a new branch locally, `pr_push()` helps you create a new pull request, and after it is merged you can use `pr_finish()` to clean things up. More about this workflow [here](https://usethis.r-lib.org/articles/pr-functions.html).