---
title: "Sliding window approach"
format: 
  html:
    toc: true
    number-sections: true
    
editor: visual
bibliography: references.bib
---

```{r setup}
#| echo: true
#| code-fold: false
#| warning: false
#| message: false

# remotes::install_github("cct-datascience/azmetr")
library(azmetr)
library(tsibble)
library(tidyverse)
library(lubridate)
library(slider)
library(here)
```

# Sliding windows

I'll explore the sliding window quantile approach used in [@faybishenko2021].
I'll use `slider` because it's tidyverse-friendly and I want to learn it.
I'll use a rolling, centered, 6-month window to calculate upper and lower 95% and 99% quantiles.
Anything outside of the rolling 99% quantile is an "extreme" value, which @faybishenko2021 assumed was bad data, and anything between the 95% and 99% quantile is an "outlier" which is suspect, but not necessarily bad.

In the plots below, I only show a few stations for clarity.

# Load Data

```{r}
targets::tar_load(daily)
```

## Mean Air Temp

```{r}
#| code-fold: true
temp_roll_test <-
  daily |> 
  select(temp_air_meanC) |> 
  group_by_key() |> 
  mutate(
    rolling_median = slide_dbl(
      temp_air_meanC,
      median,
      .before = 90,
      .after  = 90,
      .complete = TRUE
    )
  ) |>
  mutate(upper95 = slide_dbl(
    temp_air_meanC,
    \(x) quantile(x, c(0.975), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |>
  mutate(lower95 = slide_dbl(
    temp_air_meanC,
    \(x) quantile(x, c(0.025), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(upper99 = slide_dbl(
    temp_air_meanC,
    \(x) quantile(x, c(0.999), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(lower99 = slide_dbl(
    temp_air_meanC,
    \(x) quantile(x, c(0.001), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(QA = case_when(
    temp_air_meanC > lower99 & temp_air_meanC < lower95 ~ "outlier",
    temp_air_meanC > upper95 & temp_air_meanC < upper99 ~ "outlier",
    temp_air_meanC < lower99 ~ "extreme",
    temp_air_meanC > upper99 ~ "extreme",
    is.na(temp_air_meanC) ~ NA_character_
  )) 
```

```{r}
#| code-fold: true
temp_roll_test |> 
 filter(meta_station_name %in% c("Bowie", "Harquahala", "Tucson", "Maricopa")) |>
  ggplot(aes(x = datetime)) +
  geom_point(aes(y = temp_air_meanC, color = QA), size = 0.5, alpha = 0.5) +
  geom_line(aes(y = rolling_median)) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), alpha = 0.3) +
  geom_ribbon(aes(ymin = lower99, ymax = upper99), alpha = 0.3) +
  facet_wrap(~meta_station_name)
```

## Solar radiation

```{r}
#| code-fold: true

sol_roll_test <-
  daily |> 
  group_by_key() |>
  select(sol_rad_total) |>
  mutate(
    sol_rolling_median = slide_dbl(
      sol_rad_total,
      median,
      .before = 90,
      .after  = 90,
      .complete = TRUE
    )
  ) |>
  mutate(sol_upper95 = slide_dbl(
    sol_rad_total,
    \(x) quantile(x, c(0.975), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |>
  mutate(sol_lower95 = slide_dbl(
    sol_rad_total,
    \(x) quantile(x, c(0.025), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(sol_upper99 = slide_dbl(
    sol_rad_total,
    \(x) quantile(x, c(0.999), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(sol_lower99 = slide_dbl(
    sol_rad_total,
    \(x) quantile(x, c(0.001), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(QA = case_when(
    sol_rad_total > sol_lower99 & sol_rad_total < sol_lower95 ~ "outlier",
    sol_rad_total > sol_upper95 & sol_rad_total < sol_upper99 ~ "outlier",
    sol_rad_total < sol_lower99 ~ "extreme",
    sol_rad_total > sol_upper99 ~ "extreme",
    is.na(sol_rad_total) ~ NA_character_
  )) 

```

```{r}
#| code-fold: true
sol_roll_test |> 
 filter(meta_station_name %in% c("Bowie", "Harquahala", "Tucson", "Maricopa")) |>
  ggplot(aes(x = datetime)) +
  geom_point(aes(y = sol_rad_total, color = QA), size = 0.5) +
  geom_line(aes(y = sol_rolling_median)) +
  geom_ribbon(aes(ymin = sol_lower95, ymax = sol_upper95), alpha = 0.3) +
  geom_ribbon(aes(ymin = sol_lower99, ymax = sol_upper99), alpha = 0.3) +
  facet_wrap(~meta_station_name)
```

## Precipitation

```{r}
#| code-fold: true
precip_roll_test <-
  daily |> 
  select(precip_total_mm) |> 
  group_by_key() |> 
  mutate(
    precip_rolling_median = slide_dbl(
      precip_total_mm,
      median,
      .before = 90,
      .after  = 90,
      .complete = TRUE
    )
  ) |>
  mutate(precip_upper95 = slide_dbl(
    precip_total_mm,
    \(x) quantile(x, c(0.975), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |>
  mutate(precip_lower95 = slide_dbl(
    precip_total_mm,
    \(x) quantile(x, c(0.025), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(precip_upper99 = slide_dbl(
    precip_total_mm,
    \(x) quantile(x, c(0.999), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(precip_lower99 = slide_dbl(
    precip_total_mm,
    \(x) quantile(x, c(0.001), na.rm = TRUE),
    .before = 90,
    .after  = 90,
    .complete = TRUE
  )) |> 
  mutate(QA = case_when(
    precip_total_mm > precip_lower99 & precip_total_mm < precip_lower95 ~ "outlier",
    precip_total_mm > precip_upper95 & precip_total_mm < precip_upper99 ~ "outlier",
    precip_total_mm < precip_lower99 ~ "extreme",
    precip_total_mm > precip_upper99 ~ "extreme",
    is.na(precip_total_mm) ~ NA_character_
  )) 
```

```{r}
#| code-fold: true
precip_roll_test |> 
 filter(meta_station_name %in% c("Bowie", "Harquahala", "Tucson", "Maricopa")) |>
  ggplot(aes(x = datetime)) +
  geom_point(aes(y = precip_total_mm, color = QA), size = 0.5) +
  # geom_line(aes(y = precip_rolling_median)) +
  geom_ribbon(aes(ymin = precip_lower95, ymax = precip_upper95), alpha = 0.3) +
  geom_ribbon(aes(ymin = precip_lower99, ymax = precip_upper99), alpha = 0.3) +
  facet_wrap(~meta_station_name)
```

Seems a little overzealous with precip.
Likely because an emprical distribution is not really appropriate for precip data.
